---
title: "singscore"
output: html_document
date: '2022-06-23'
---

# Preparation of data

```{r data_loading}
## Load data

library(tidyverse)

# Load counts
counts.original <- read.delim("C:/Users/paucl/Documents/TFM/dades/SERRAVIO_03_04_ProcessatCNAG/COUNTS_by_genes/COUNTS_genes_SERRAVIO_03_04", header=T)

# Load phenodata
phenodata.original <-as.data.frame( readxl::read_xlsx("C:/Users/paucl/Documents/TFM/dades/20220406SamplesInfo_VS.xlsx", sheet=1, col_names = T, trim_ws=T))

# Change phenodata variables names
colnames(phenodata.original)[colnames(phenodata.original)=="Response (R/S)"] <- "Response"
colnames(phenodata.original)[colnames(phenodata.original)=="PDX name"] <- "PDX_name"
```

```{r modifications_on_data0} 
# Create phenodata variable Original PDX
phenodata.original$Original_PDX <- unlist(strsplit(phenodata.original$PDX_name, "OR"))

# Set Original_PDX from PDX474.7 as PDX474.1 as it is the same tumour after adquiring resistance in patient
phenodata.original$Original_PDX[grep("PDX474.7",phenodata.original$Original_PDX)] <- "PDX474.1"

# Set variables as factors in phenodata and set a reference level
cols <- c("Clinical subtype","Tumor origin","Patient ID","Alteration in HRR pathway", "Drug", "Response", "Original_PDX")
phenodata.original[,cols] <- lapply(phenodata.original[cols], factor)

# Change reference level
phenodata.original$Drug <- relevel(phenodata.original$Drug, ref="control")
phenodata.original$Response <- relevel(phenodata.original$Response, ref="Sensitive")
```

First column, containing gene metadata is splitted into ensembl_id, gene_symbol and gene_product type. Then mouse genes are removed and remaining genes are filtered by protein_coding genes. Afterwards a separate table with gene metadata is created from the information in counts table (which now contains only counts).

Samples match between phenodata and count table is checked (and in the same order)

```{r gene_metadata_creation}
# Split gene information from first column in counts
counts.original <- cbind(str_split_fixed(counts.original[,1], ",", n=3), counts.original)[,-4]

# Add names to the new columns
colnames(counts.original)[1:3] <- c("ensembl_id_version", "gene_symbol", "gene_product")

# Remove human genes
human_genes_index <- unlist(lapply(strsplit(counts.original$ensembl, "M"), "[[",1))=="ENS"
counts.original <- counts.original[human_genes_index,]

# Select protein_coding genes
counts <- counts.original %>%
  dplyr::filter(gene_product=="protein_coding")

# Assign ensembl ID as rownames
rownames(counts) <- counts[,1]

# Create separate dataframe with gene metadata
gene_metadata <- counts[,1:3]

# Remove gene metadata from counts table
counts <- counts[,-1:-3]

# Arrange by model name, so both tables are in the same sample order
phenodata.original <- arrange(phenodata.original, `Sample Barcode`)[1:68,]
counts <- counts[,order(colnames(counts))]

# Check that all samples in count table are in phenodata
sum(colnames(counts) %in% phenodata.original[,2])

# Check that samples are in the same order
sum(colnames(counts) == phenodata.original[,2])/dim(phenodata.original)[1]
```

```{r subsetting}
phenodata <- phenodata.original

# Select only this mouse strain (the other only has macrophages)
counts <- counts[,phenodata$`Mouse strain`=="Hsd:Athymic Nude-Foxn1nu"]

phenodata <- phenodata%>%
  mutate(Group=as.factor(paste(Response, Drug, sep="_")))%>%
  filter(`Mouse strain`=="Hsd:Athymic Nude-Foxn1nu")
```

```{r new_column_names}
# Change names of samples
new.names <- c()
for(sample in colnames(counts)){
  new.names <- c(new.names,paste0(phenodata.original[phenodata.original[,2]==sample,4], "_",phenodata.original[phenodata.original[,2]==sample,"Drug"]))}

# Assign the names
colnames(counts) <- new.names
rownames(phenodata) <- new.names

# Save number of samples
nsamples <- nrow(phenodata)
```

```{r}
# Check that there are no NA
non_na <- complete.cases(counts)
sum(!non_na)# should be 0

# Subset 
counts <- counts[non_na,]
```

```{r}
# Create a column with ensembl_id
gene_metadata <- gene_metadata %>%
  filter(ensembl_id_version %in% rownames(counts))%>%
  mutate(ensembl_id= gsub("\\..*$", "",ensembl_id_version))
```

```{r}
library(biomaRt)
# Find human ortholog ENSEMBLID 
mart1 = useEnsembl("ensembl", dataset="hsapiens_gene_ensembl", host = "https://dec2021.archive.ensembl.org")
mart2 = useEnsembl("ensembl", dataset="mmusculus_gene_ensembl", host = "https://dec2021.archive.ensembl.org") 

# Trasnformation human / mouse
human_ensembl <- getLDS(attributes=c("ensembl_gene_id"),
                        filters="ensembl_gene_id", 
                        values=gene_metadata$ensembl_id, 
                        mart=mart2,
                        attributesL=c("ensembl_gene_id"),
                        martL=mart1)

# remove duplicated mouse genes
human_ensembl <-human_ensembl%>%
  distinct(Gene.stable.ID, .keep_all=T)

# add human ensembl to gene_metadata
gene_metadata <- gene_metadata%>%
  right_join(human_ensembl, by=c("ensembl_id"="Gene.stable.ID"))


# remove counts with unknown human gene
counts <- counts%>%
  filter(rownames(counts)%in%gene_metadata$ensembl_id_version)

#mirar llargada i mirar si conservem els gens de la signatura.
```

```{r}
# Download gene start and end positions to compute gene length 
mouse <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")
res <- getBM(attributes = c("ensembl_gene_id",'transcript_start','transcript_end'), 
             mart = mouse, 
             filter= "ensembl_gene_id", 
             values = gene_metadata$ensembl_id)
# COmpute gene_length
genelengths <- res%>%
  mutate(gene_length=transcript_end-transcript_start)%>%
  group_by(ensembl_gene_id)%>%
  summarise_at(vars(gene_length) , list(name = mean))
  

gene_metadata <- gene_metadata%>%
  right_join(genelengths, by=c("ensembl_id"="ensembl_gene_id"))
```

```{r}
## Data exploration
# Library size
lib.size <- colSums(counts, na.rm=T) #Total number of counts of each sample

# Check distributions by response
boxplot(lib.size~as.factor(phenodata$"Group"),col=as.factor(phenodata$"Group"))

par(mar=c(10,5,2,2)+.1)
barplot(lib.size, las=2, col=as.factor(phenodata$"Group"))
legend("topleft", levels(phenodata$Group), fill=1:4)
title("Library size per sample by response")

# Remove the sample with much bigger library
counts <- counts[lib.size!=max(lib.size)]
phenodata <- phenodata[lib.size!=max(lib.size),]
```

```{r}
# Compute TMM
TMM <- tweeDEseq::normalizeCounts(counts, method="TMM")
rownames(TMM) <- toupper(gene_metadata$gene_symbol)
TMM <- as.data.frame(TMM)

# Compute TPM
rpk <- apply(counts, 2, 
             function(x) x/(gene_metadata$name/10^3)) #nameis the length of the gene
# Normalize gene-normalized values by the sample size using rpk values
NormByTPM <- apply(rpk, 2, 
                   function(x) x / sum(as.numeric(x), na.rm=TRUE) * 10^6)
colSums(NormByTPM, na.rm=TRUE)#equals 1e+06

# Set symbol as rowname
rownames(NormByTPM) <- toupper(gene_metadata$gene_symbol)
NormByTPM <- as.data.frame(NormByTPM)
```


# Application of Deconvolution Methods

```{r}
# Load Packages
packages <- c("dplyr", "stringr", "ggplot2", "biomaRt", "pheatmap", "devtools")
lapply(packages, library, character.only = TRUE)

```

```{r}
## EPIC (TRef reference-based method)

#install_github("GfellerLab/EPIC", build_vignettes=TRUE)
library(EPIC)

# Run EPIC cell type deconvolution analysis
out <- EPIC(bulk = as.matrix(NormByTPM))

# Add group factor and pivot longer to input in ggplot
cellfractions <- as.data.frame(out$cellFractions)

cellfractions$sample <- rownames(out$cellFractions)
cellfractions$group <- phenodata$Group

cellFrac <- tidyr::pivot_longer(as.data.frame(cellfractions), 1:8, "cell")

# Barplot by sample filled by cell type
ggplot(cellFrac)+
  geom_col(aes(x = sample, y = value, fill = cell))

# Boxplot dodged by group
ggplot(cellFrac)+
  geom_boxplot(aes(x = group, y = value, position = "dodge", fill = cell), trim=TRUE)+
  theme_classic()

# Boxplot of cell types by group
for(celltype in levels(as.factor(cellFrac$cell))){
print(ggplot(cellFrac %>%
         filter(cell == celltype), aes(x = group, y = value, fill = cell))+
  geom_boxplot()  +
  theme_classic() +
  ggtitle(celltype))}
```

```{r}
## CIBERSORT (lm22 reference-based method)

#install_github("IOBR/IOBR")
library(IOBR)

# Run cibersort
cibersort <- CIBERSORT(sig_matrix = lm22, mixture_file = as.data.frame(NormByTPM), perm = 1000, QN = T, absolute = F) # absolute=T, abs_method = "sig.score"

# Add columns with conditions info
cibersort <- as.data.frame(cibersort)
cibersort$group <- phenodata$Group

# Check distribution of correlations by group
boxplot(cibersort$Correlation~cibersort$group)

# Pivot data
ciberlong <- tidyr::pivot_longer(as.data.frame(cibersort), 1:25, "cell" )
ciberlong$cell <- as.factor(ciberlong$cell)

# Remove unwanted info for plotting
ciberlong <- ciberlong%>%
  filter(!(cell%in%c("P-value","Correlation","RMSE")))

# Boxplot of cell types by group
for(celltype in levels(as.factor(ciberlong$cell))){
print(ggplot(ciberlong %>%
         filter(cell == celltype), aes(x = group, y = value, fill = cell))+
  geom_boxplot()  +
  theme_classic() +
  ggtitle(celltype))}
```

```{r}
## MCPcounter (Marker-based method)

#install_github("ebecht/MCPcounter", ref="master", subdir="Source")
library(MCPcounter)

# Run MCPcounter
mcpres <- MCPcounter.estimate(NormByTPM,
                    featuresType="HUGO_symbols")

# Add group info
mcpres <- as.data.frame(t(mcpres))
mcpres$group <-phenodata$Group
mcpres$sample <- rownames(mcpres)

# Pivot data
mcpreslong <- tidyr::pivot_longer(mcpres, 1:9, "cell")


# Boxplot of cell types by group
for(celltype in levels(as.factor(mcpreslong$cell))){
print(ggplot(mcpreslong %>%
         filter(cell == celltype), aes(x = group, y = value, fill = cell))+
  geom_boxplot()  +
  theme_classic() +
  ggtitle(celltype))}

# Heatmap
column_ha = HeatmapAnnotation(Group = phenodata$Group)
Heatmap(t(mcpres[1:7]), top_annotation=column_ha)
```


```{r}
## quanTIseq (TIL10 Reference-based method) (probably the best so far)
library(quantiseqr)

# Run quanTIseq
quantires <- quantiseqr::run_quantiseq(NormByTPM,
               signature_matrix=,
               is_arraydata=FALSE,
               is_tumordata=TRUE,
               scale_mRNA=TRUE)

# Add group info and pivot
quantires$group <- phenodata$Group
quantilong <- tidyr::pivot_longer(quantires, 2:12, "cell" )

# Quantiseq plot
quantiplot(quantires)

# Boxplot of cell types by group
for(celltype in levels(as.factor(quantilong$cell))){
print(ggplot(quantilong %>%
         filter(cell == celltype), aes(x = group, y = value, fill = cell))+
  geom_boxplot()  +
  theme_classic() +
  ggtitle(celltype))}
```

```{r}
# TIMER (reference-based method)
timer <- immunedeconv::deconvolute_timer(TMM, indications = rep("brca", dim(TMM)[2]))

# Add group info and pivot data
timer <- as.data.frame(t(timer))
timer$group <- phenodata$Group

timerlong <- pivot_longer(timer, 1:6, "cell")

# Boxplot of cell types by group
for(celltype in levels(as.factor(quantilong$cell))){
print(ggplot(quantilong %>%
         filter(cell == celltype), aes(x = group, y = value, fill = cell))+
  geom_boxplot()  +
  theme_classic() +
  ggtitle(celltype))}
``` 




```{r}
# Singscore (rank-based single sample scoring method)
library(singscore)
library(GSEABase)
library(SummarizedExperiment)

# rank genes
rankData <- rankGenes(as.matrix(TMM))
# scoring
nk <- readxl::read_xlsx("C:/Users/paucl/Documents/TFM/cell type deconvolution/NKsignature.xlsx")%>%
  filter(`Final list`==TRUE)
NKsig <- as.vector(unlist(nk[1]))

# check for genes in signature
sum(rownames(NormByTPM)%in%NKsig)

# perform score
scoredf <- simpleScore(rankData, upSet= NKsig)

# plot ranks
plotRankDensity(rankData[,2,drop = FALSE], upSet =NKsig , isInteractive = FALSE)

scoredf$group <- phenodata$Group

# Barplot
ggplot(scoredf)+
  geom_bar(aes(x=rownames(scoredf),y=TotalScore, fill=phenodata$Group), stat="identity")

# Boxplot by groups
ggplot(scoredf)+
  geom_boxplot(aes(x=group,y=TotalScore, fill=group))


barplot(scoredf$TotalScore)

# Permutation test
ncores <- 1

permuteResult <-
  generateNull(
    upSet = NKsig,
    rankData = rankData,
    subSamples = 1:63,
    centerScore = TRUE,
    knownDirection = TRUE,
    B = 1000,
    ncores = ncores,
    seed = 1,
    useBPPARAM = NULL
  )

# get pvals
pvals <- getPvals(permuteResult, scoredf, subSamples =1:63)


plotNull(permuteResult, scoredf, pvals, sampleNames = names(pvals)[1])

pvals <- as.data.frame(pvals)
pvals$group <- phenodata$Group

ggplot(as.data.frame(pvals))+
  geom_density(aes(x=pvals))
  
ggplot(as.data.frame(pvals))+
  geom_boxplot(aes(x=group, y=pvals, fill=group))
```

```{r}
library(ComplexHeatmap)
nkgep <- NormByTPM%>%
  filter(rownames(NormByTPM)%in%NKsig)

pheatmap::pheatmap(as.matrix(nkgep), scale="row")



column_ha = HeatmapAnnotation(foo1 = phenodata$Group)
nkgepscaled <-pheatmap:::scale_rows(nkgep)

#with clustering
Heatmap(nkgepscaled, top_annotation=column_ha)

#without clustering
Heatmap(nkgepscaled, top_annotation=column_ha, column_order= rownames(phenodata%>%dplyr::arrange(by_group=Group)))
```