---
title: "Untitled"
output: html_document
date: '2022-06-29'
---



```{r}
library(tidyverse)
library(ggrepel)
library(ggpubr)
library(ggpmisc)

# load top tables
sensitives <- read.csv("C:/Users/paucl/Documents/TFM/DEA/results/drug_sensitive/toptable_olaparib_vs_control_allsensitivenoisyrdensity.csv")
resistants <- read.csv("C:/Users/paucl/Documents/TFM/DEA/results/drug_resistant/toptable_olaparib_vs_control_allresistantnoisyrdensity.csv") 

# load hallmark
hallmark <- fgsea::gmtPathways("C:/Users/paucl/Documents/TFM/dades/collectionsets_msigdb/h.all.v7.5.1.symbols.gmt")

# load nice names for hallmark
nice_names <- read.csv("C:/Users/paucl/Documents/TFM/dades/collectionsets_msigdb/beautiful_names.csv")
names(hallmark) <- as.vector(nice_names[,2])

# join tables
all <- left_join(sensitives, resistants, by="symbol", suffix=c(".s",".r"))
```


```{r}
# compute mean of logFC by geneset
inset <- lapply(hallmark , 
            \(x) lapply(subset(all ,  all$symbol %in% x)[grep("log2FoldChange",
                                                              colnames(all))],
            mean))

meanlogFCin <- do.call(rbind.data.frame, inset)

# compute mean of logFC by geneset for genes not in the set
outset <- lapply(hallmark , 
            \(x) lapply(subset(all ,  !(all$symbol %in% x))[grep("log2FoldChange",
                                                              colnames(all))],
            mean))

meanlogFCout <- do.call(rbind.data.frame, outset)

# join inset and outset
meanlogFC <- left_join(add_rownames(meanlogFCin), add_rownames(meanlogFCout), suffix=c(".in", ".out"), by=c("rowname"))

meanlogFC <- as.data.frame(meanlogFC)

# add number of genes in the set
meanlogFC$`Genes in set` <- unlist(lapply(hallmark, length))


# compute difference between in and outs
meanlogFC <- meanlogFC %>%
  mutate(sensitive= log2FoldChange.s.in-log2FoldChange.s.out)%>%
  mutate(resistant= log2FoldChange.r.in-log2FoldChange.r.out)


```


Comparison of olaparib-treatment effect on sensitive and resistant PDX of average log2 fold change of gene set member minus non-member genes for every gene set in MsigDb Hallmark collection.

Significance has been obtained from a two-sided Wilcoxon Rank-Sum test on the differences between log2 fold changes from gene set member genes from resistant and sensitive contrasts, after rejecting the null hypothesis in Shapiro-Wilk normality test.

The lineal model built with all data points has a slope = 0.45 with an adjusted R-squared of 0.87.

All significant gene sets suffer a greater effect in sensitive PDX as all are true `(abs(meanlogFC$log2FoldChange.s.in)-abs(meanlogFC$log2FoldChange.r.in))[meanlogFC$wilcox.p<0.05]`

Green - there are differences between logFC between sensitives and resistants
QI over line -> higher upregulation in resistant
QI under line -> higher upregulation in sensitive

QIII over line -> higher downregulation in sensitive
QIII under line -> higher downregulation in resistant

QIV -> sensitive upregulation but resistant downregulation, which one has the biggest effect depend on the value (a line with -1 slope ccould make a separation)

```{r}
insetlogs <- lapply(hallmark , 
            \(x) subset(all ,  all$symbol %in% x)[grep("log2FoldChange",
                                                              colnames(all))])

logFCin <- do.call(rbind.data.frame, insetlogs)
logFCin$set <- gsub("\\..*", "",rownames(logFCin))

tapply(logFCin$log2FoldChange.s-logFCin$log2FoldChange.r, factor(logFCin$set), shapiro.test)

wilcox.res.log <- tapply(logFCin$log2FoldChange.s-logFCin$log2FoldChange.r, factor(logFCin$set), wilcox.test, alternative="two.sided", mu=0)

wilcox.p <- unlist(lapply(wilcox.res.log, "[[", 3))
meanlogFC$wilcox.p <- wilcox.p[meanlogFC$rowname]

meanlogFC$wilcox.p.signif <- ifelse(meanlogFC$wilcox.p < 0.05,  "< 0.05", "ns")


boxplot(logFCin$log2FoldChange.s- logFCin$log2FoldChange.r~ factor(logFCin$set))
abline(h=0)

meanlogFC[!((meanlogFC$log2FoldChange.s.in>0 & meanlogFC$log2FoldChange.r.in>0)|(meanlogFC$log2FoldChange.s.in<0 & meanlogFC$log2FoldChange.r.in<0)), "rowname"]

#write.csv(meanlogFC, "C:/Users/paucl/Documents/TFM/DEA/results/bidimensional/bidimensionalLogFCgeneset.csv")
```

```{r}
# Load GSEA results
contrast.list <- c("drug_sensitive", "drug_resistant", "resistance_control", "resistance_olaparib")
contrast_names <- c("olaparib_vs_control_allsensitive", "olaparib_vs_control_allresistant", "sensitive_vs_resistant_allcontrol", "sensitive_vs_resistant_alltreated")
names(contrast_names) <- contrast.list

for(contrast in contrast.list){
  assign(paste0("GSEA.", contrast),
         read.csv(paste0("C:/Users/paucl/Documents/TFM/DEA/results/",
                          contrast, "/GSEAres_",
                          contrast_names[contrast],"noisyrdensity.csv")))}
meanlogFC <- meanlogFC[order(meanlogFC$rowname),]
meanlogFC$GSEA <- c(ifelse(GSEA.drug_sensitive[order(GSEA.drug_sensitive$pathway), "padj"] < 0.05, 
                         ifelse(GSEA.drug_resistant[order(GSEA.drug_resistant$pathway),
                                                    "padj"] < 0.05, 
                                "Sensitive & Resistant",
                                "Sensitive"), 
                         ifelse(GSEA.drug_resistant[order(GSEA.drug_resistant$pathway),
                                                    "padj"] < 0.05, 
                                "Resistant", 
                                "None")))

meanlogFC$plotnames <- ifelse(meanlogFC$wilcox.p.signif == "< 0.05", meanlogFC$rowname, "")

```


```{r}
# Two contrasts plot
ggplot(meanlogFC, 
       aes(x = sensitive, 
           y = resistant,
           label = plotnames))+
  geom_hline(yintercept=0, color="darkgrey", linetype="dashed") +
  geom_vline(xintercept=0, color="darkgrey", linetype="dashed") +
  geom_abline(slope=1, size=0.3, color="black", linetype="dotted", inherit.aes=F)+
  stat_smooth(method = "lm", se=F, color="black", linetype="twodash", size= 1.3)+
  geom_point(aes(size =`Genes in set`,
                 fill = GSEA,
                 color = wilcox.p.signif),
             alpha=0.7,
             shape=21,
             stroke= 1.2)+
  # add text
  geom_text_repel(box.padding = 0.5, 
                  max.overlaps = Inf,
                  size=3)+
  scale_fill_manual(values = c("Sensitive & Resistant" = "#4b9600",
                              "Sensitive" = "#1984c5",
                              "Resistant" = "#C309C7",
                              "None" = "darkgrey"),
                     breaks = c("Sensitive", "Resistant", "Sensitive & Resistant", "None"))+
  scale_color_manual(values= c("< 0.05" = 17,
                               "ns" = 16))+
  guides(fill = guide_legend(override.aes = list(size=5),
                               title = "GSEA FDR <0.05"),
         color = guide_legend(override.aes = list(size=5),
                              title= "Treatment effect \nby response"))+
  theme_classic()+
  scale_size_area()+
  labs(x = bquote(bold(.(log[2] ~ FC ~Treated~ vs ~Control~ (Sensitive)))), 
       y = bquote(bold(.(log[2] ~ FC ~Treated~ vs ~Control~ (Resistant)))),
           title = "")+
  theme(axis.line = element_line(size=0.7),
        axis.ticks = element_line(size=0.7),
        axis.text = element_text(face="bold"),
        axis.title = element_text(face="bold"),
        legend.box = "vertical")+
  stat_regline_equation(aes(label= after_stat(eq.label)),
                        data = meanlogFC,
                        formula=lm(meanlogFC$resistant ~ meanlogFC$sensitive), 
                        label.x.= 0.8 , 
                        label.y=0.1)+
  stat_poly_eq(aes(label= after_stat(adj.rr.label)), label.x= 0.9, label.y = 0.55)+
  coord_fixed(ratio=1)
ggsave("C:/Users/paucl/Documents/TFM/DEA/results/bidimensional/bidimensionalGSEAnew_noisyrdensity.png", units="in", limitsize = FALSE, width=12, height=5)
```





```{r}
# All genes bicontrast plot

all$signif <- ifelse(all$padj.s < 0.05, ifelse(all$padj.r<0.05, "Sensitive & Resistant", "Sensitive"), ifelse(all$padj.r<0.05, "Resistant", "None"))

suball <- base::subset(all, all$padj.s<0.05 | all$padj.r<0.05)

ggplot(all)+
  geom_point(aes(x = log2FoldChange.s, 
                 y = log2FoldChange.r, 
                 color = signif),
             alpha=0.4)+
  scale_color_manual(values = c("Sensitive & Resistant" = "Red",
                                "Sensitive" = "Blue",
                                "Resistant" = "Green",
                                "None" = "darkgrey"))+
  theme_classic()+
  geom_hline(yintercept=0, color="darkgrey", linetype="dotted") +
  geom_vline(xintercept=0, color="darkgrey", linetype="dotted") +
  geom_abline(slope=1, size=1, color="grey", linetype="dashed", inherit.aes=F)+
  scale_size_area()+
  labs(x = bquote(bold(.(log[2] ~ FC ~ Sensitive ~(olaparib~ vs ~control)))), 
       y = bquote(bold(.(log[2] ~ FC ~ Resistant ~(olaparib~ vs ~control)))),
           title = "")+
  # add text
  geom_text_repel(data= suball,
                  aes(x = log2FoldChange.s,
                      y = log2FoldChange.r,
                      label = symbol),
                   size = 3,
                   min.segment.length = 0.3)+
  guides(color= guide_legend("FDR<0.05"))+
  theme(axis.line = element_line(size=0.7),
        axis.ticks = element_line(size=0.7),
        axis.text = element_text(face="bold"),
        axis.title = element_text(face="bold"))
#ggsave("C:/Users/paucl/Documents/TFM/DEA/results/bidimensional/bidimensionalLogFC.png", units="in", limitsize = FALSE, width=10, height=6)
```






